TIMES Ireland Model: Residential Sector

### 2021-05-20 manuscript.tex
20May2021

\subsection{Ireland: Residential Sector}
\label{Intro:IreRSD}
\par

The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting. Under these guidelines, Ireland’s residential sector accounted for 14.2\%of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Focusing on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions. Fig. \ref{fig:Ireland2018GHG}. shows the energy related GHG disaggregated between transport, heat and electricity, where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in Ireland energy decarbonisation.

 \par The residential sector target is expected to be reduce GHG emissions beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with other sectors. Despite these high target ambitions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions. The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity. The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}. Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity and heat, but in transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.

\par

The CAP2019 Marginal Abatement Cost Curve (MACC) developed by McKinsey and Company's modelling tools is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} and provides an analytical foundation on the most cost-effective pathway to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. The CAP2019 MACC in Fig.\ref{fig:CAP2019:MACC} explores technological, behavioural and fuel switching options. The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$. The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$. over the 2021 to 2030 period. The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. For example, increasing onshore-wind by $8.2 GW$ and offshore-wind would grow to $3.5 GW$,this would be expected to save $7 to 8 MtCO2_e_q$. This would help to indirectly reduce the GHG in the residential sector. Solutions specific to the residential sector are: 
\begin{itemize}
    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
    \item{Switch from oil boilers to heat pumps in existing dwellings}
    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
\end{itemize}

Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually building due to grants, awareness and increasing the ease of upgrades for homeowners. In parallel, new homes have a strict regulation to comply with, which means despite the high capital costs heat pumps are the preferred heating technology. The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}. The CAP2019 plan to deliver a new retrofitting model to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlighting the energy efficiency first approach in how Ireland aims to progress.

@@

\par
Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s high reliance on fossil fuels, larger dwellings and rural settlement patterns of low thermal efficiency of the housing stock. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, and 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\%. \cite{EurostatShareSurveyilc_lvho50a}. The lower population density and large dwellings of the existing Irish housing stock, translates to more expensive GHG mitigation options compared to other residential sectors in the EU.

@@

%PICK UP HERE
Historical fuel usage in one part of understanding residential energy, the other main part is understanding the low thermal efficiency of Irelands housing stock. As mentioned in section \ref{Intro:IreRSD}, Irelands new buildings produce 70\% less GHG emissions than building in 2005, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock highlights the impact the year of construction has on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}.The BERs are categorised into BER labels, with A1 representing the lowest energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest energy consumption ($ >450 kWh/m^2/year$) .

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.4]{Figures/BERAge.png}
 \caption{Building Energy Rating by Age}
 \label{fig:BERAge}
\end{figure}

The poor BERs consequently translates to higher energy bills to maintain comfortable temperature, but in reality it relates to low internal temperatures, which likely results in higher excessive winter mortality \cite{Clinch2000HousingMortality}.The BERs which come from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock. The BERs are required for new houses, rented houses, sold houses and renovated houses. However, BERs do not take account of actual usage, i.e occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, I will explain how I approached this problem in the Literature review.

Fuel Poverty is when a household spends \textit{``10\% or more of income to achieve adequate warmth"}, an element energy reported 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. While income supports and the \textit{Fuel Allowance and the Household Benefits} provides support to low income households to allievate some the fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. As Kerr points out the link between energy and social welfare policy is made explicitly \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}. Ironically, this places low income households deeply dependent on fossil fuels for health.

@@

Space heating accounted for the majority of final energy, Fig. \ref{fig:Ireland2018ServiceDemands} shows a snapshot of the residential energy services demands in 2018. As households become more thermal efficient and use higher efficient heating, space heating energy would be expected to decline. The new A-rated dwellings will use less energy for space heating, but Pumps \& Fans energy will increase as mechanical ventilation is used over natural ventilation.

@@

Space heating is therefore central to residential energy policy, other important service energy demands are shown in Fig.\ref{fig:Ireland2018ServiceDemands}. Notably space cooling is absent for the pie chart, but space cooling is accounted for in appliances. The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output space cooling energy consumption data, shown in Fig.\ref{fig:Cooling} a significant amount of space cooling energy is expected in the Ireland residential sector.

@@

\subsection{Current Energy Policy}
\label{Intro:Policy}
Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}. The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}.
The 2019 Climate Action Plan identified 29 actions that would reduce emissions from the sector by 40-45%.  These actions included:
Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings
500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent
Installation of 600,000 heat pumps, with 400,000 of these in existing buildings
Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes
Complete the rollout of the Support Scheme for Renewable Heat
Increasing the number of Sustainable Energy Communities to 1,500

\subsubsection{Electricity}
\par
Decarbonising Electricity is key for Ireland, both the \textit{``stick"} (tax) approach which aims to deter investment in fossil fuel power generation, and the \textit{``carrot"} (subsidies) approach which aims to encourage renewable electricity is applied in Irish policy.
In Ireland, a €15/tCO2 carbon tax was introduced in 2009 to influence fossil fuel demand. The carbon tax receipts grew every year from 2010 to 2018 \cite{AnOffice}. Ireland's current carbon tax of €33.50/tCO2 will rise to €100/tCO2 in 2030. Many countries have taken this approach, especially in Europe, where Finland was the first country to introduce a carbon tax in 1990, while Sweden were second to introduce a carbon tax after Finland, the carbon tax rate is the highest in the world at €108.81/tCO2. A carbon price or tax now accounts for approx. 15\% of GHG emissions \cite{WorldBankGroup2020State2020}. 
The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentives investment in renewable electricity to achieve legally binding renewable targets under 2009/28/EC. REFIT operates  by  guaranteeing  new renewable generation and biomass a minimum price for electricity over a 15 year period. The last REFIT schemes closed in December 2015. In 2020 REFIT was replaced by the Renewable Electricity Support Scheme (RESS), which supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. RESS delivers for communities to participate in renewable energy projects and increasing technology diversity by broadening the renewable electricity technology mix, where almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this builds alot on the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW by 2030.

\subsubsection{Heat}
Domestic supports for renewable heat or energy efficiency is through SEAI grants,\begin{itemize}
    \item {Insulation}
    \subitem {Attic}
    \subitem {Cavity Wall}
    \subitem {Internal - Dry Lining}
    \subitem {External}
    \item {Heat Pumps}
    \subitem {Air to Water}
    \subitem {Ground Source to Water}
    \subitem {Exhaust Air to Water}
    \subitem {Water to Water}
    \subitem {Air to Air}
    \item {Heating Controls}
    \item {Solar Water Heating}
\end{itemize}

This report is a study of potential design options for a Renewable Heat Incentive (RHI) f

>>>>

@@

\par
Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\%. \cite{EurostatShareSurveyilc_lvho50a}. The low building thermal efficiency is highlighted in Fig.\ref{fig:BERAge} and home heating oil accounts for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport} .

@@

The building regulations in 2021 will require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock. The BERs are required for new houses, rented houses, sold houses and renovated houses.

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.5]{Figures/BERAge.png}
 \caption{Building Energy Rating by Age}
 \label{fig:BERAge}
\end{figure}

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which results in higher excessive winter mortality \cite{Clinch2000HousingMortality}. BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, I will explain how I approached this problem in the literature review.

Fuel Poverty which is a major contributor to the high excessive winter deaths, is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"}. An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. As Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty, which include increased risk of catastrophic cardiovascular or cerebrovascular events \cite{OMeara2016AIreland}. Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.

The United Nations Framework Convention on Climate Change (UNFCCC) GHG reporting rules, as per Decision 24/CP.19, states that 2006  Intergovernmental Panel on Climate Change (IPCC) guidelines are to be adhere to for national GHG reporting. Under these guidelines, Ireland’s residential sector accounted for 14.2\%of GHGs emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Focusing on energy-related GHG emissions, the residential sector directly accounts for 24\% of GHG emissions. Fig. \ref{fig:Ireland2018GHG} shows the energy related GHG disaggregated between transport, heat and electricity. Where the residential sector is the direct source 47\% of heat and 30\% of electricity GHG emissions, the residential sector is therefore a pivotal sector in Ireland energy decarbonisation.

@@

Space heating is the largest residential energy service demand, accounted for 61.8\% of final energy usage. Other residential energy services and their share of final energy in 2018 is shown in \ref{fig:Ireland2018ServiceDemands}. As building thermal efficiency and heating technology efficiency improves, both space and water heating energy would be expected to decline. The new A-rated dwellings can be used to predict the share of future energy service demands. From the BER database, it can be seen that A-rated dwellings use less than half space heating energy of the average dwelling and A-rated water heating energy decreases by about 40\% compared to the average A-rated dwelling. However Pumps \& Fans energy increases as mechanical ventilation is used over natural ventilation and Lighting usage in A-rated is more than double the average household. Cooking and Appliances energy usage would be expected to remain stable.

@@

Notably space cooling is absent from Fig \ref{fig:Ireland2018ServiceDemands}, but space cooling is accounted for in appliances. The Policy Oriented Tool for Energy and Climate Change Impact Assessment (POTEnCIA) project disaggregated Ireland's residential space cooling from appliances and have output space cooling energy consumption data, shown in Fig.\ref{fig:Cooling} a significant increase in space cooling energy is expected in the Ireland residential sector.

@@

\subsection{Current Energy Policy}
\label{Intro:Policy}

The energy-related GHG emissions is expected to be reduce GHG emissions beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture. Despite these high target ambitions in the energy-related GHG emissions, preliminary results from the Environmental Protection Agency (EPA) has indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions. The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity & transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. The future decarbonisation of the residential sector is dependant on receiving low carbon intensity electricity.

Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}. The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}. Some of these actions included:
\begin{itemize}
    \item{Reducing fossil fuel use in the sector and moving to electricity to provide heating and hot water in buildings}
    \item{500,000 building retrofits to a BER B2 / cost optimal equivalent or carbon equivalent}
    \item{Installation of 600,000 heat pumps, with 400,000 of these in existing buildings}
    \item{Development of district heating, including two initiatives of municipal scale with the potential to provide heat equivalent to the needs of about 50,000 homes}
    \item{Complete the roll-out of the Support Scheme for Renewable Heat}
    \item{Increasing the number of Sustainable Energy Communities to 1,500}
\end{itemize}

\par
The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_e_q$. The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_e_q$. over the 2021 to 2030 period. The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_e_q$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. The options specific to the residential sector are: 
\begin{itemize}
    \item{Retrofit oil boiler existing dwellings to B2 equivalent}
    \item{Switch from oil boilers to heat pumps in existing dwellings}
    \item{Retrofit gas boiler and solid fuel stove existing dwellings to B2 equivalent}
    \item{Introduce CO2-free heating in new buildings}\cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
\end{itemize}

Where, the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie. save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie. cost money) but the expected abatement of carbon is far greater.

\subsubsection{Electricity}
\par
The decarbonisation of the power generation sector is well underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}. Under CAP2019 the continued growth and decarbonisation of the electricity will be the fundamental to lowering energy-related GHG in not only electricity, but in heat and transport also \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
Decarbonising electricity is key for Ireland, both the \textit{``stick"} (tax) approach which aims to deter investment in fossil fuel power generation, and the \textit{``carrot"} (subsidies) approach which aims to encourage renewable electricity is applied in Irish policy.
In Ireland, a €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce fossil fuel demand and investment. The carbon tax receipts grew every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. Many countries have taken this approach, especially in Europe, where Sweden were second to introduce a carbon tax after Finland, the Swedish carbon tax rate is the highest in the world at €108.81/ $tCO_2$.  About 15\% of global GHG emissions have either an associated cost or tax, this figure is increasing incrementally \cite{WorldBankGroup2020State2020}. 
The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise Ireland to invest in renewable electricity to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing  new renewable generation and biomass a minimum price for electricity over a 15 year period. The last REFIT schemes closed in December 2015. In 2020 REFIT was replaced by the Renewable Electricity Support Scheme (RESS), which supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. RESS delivers for communities to participate in renewable energy projects and increasing technology diversity by broadening the renewable electricity technology mix, where almost 800 MW of solar has been approved in RESS-1 \cite{Eirgrid2020RenewableResults}, this builds alot on the current solar capacity of 16 MW, while making significant progress towards the CAP2019 target of 1500 MW by 2030.

\subsubsection{Heat}
Sustainable Energy Authority of Ireland (SEAI) and others have been part of the renovation wave in Ireland, the uptake of heat pumps in existing homes is gradually growing due to grants, awareness and ease of upgrades for homeowners. In parallel, new homes have a strict regulation to comply with, which means despite the high capital costs heat pumps are the preferred heating technology. As mentioned in Section\ref{Intro:IreRSDBack} new buildings in Ireland produce 70\% less GHG emissions than building in 2005 and Section\ref{Intro:Policy} mentioned plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030.  The energy efficiency first approach in how Ireland aims to 
decarbonize the residential sector. Domestic supports for renewable heat or energy efficiency include:
\begin{itemize}
    \item {Insulation}
    \subitem {Attic}
    \subitem {Cavity Wall}
    \subitem {Internal - Dry Lining}
    \subitem {External}
    \item {Heat Pumps}
    \subitem {Air to Water}
    \subitem {Ground Source to Water}
    \subitem {Exhaust Air to Water}
    \subitem {Water to Water}
    \subitem {Air to Air}
    \item {Heating Controls}
    \item {Solar Water Heating}
\end{itemize}
%PICKUPHERE
This report is a study of potential design options for a Renewable Heat Incentive (RHI)
